---
title: 'SHM: Quantifing selection pressures'
author: "Namita Gupta"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_height: 4
    fig_width: 7.5
    highlight: pygments
    theme: readable
    toc: yes
  pdf_document:
    fig_height: 4
    fig_width: 7.5
    highlight: pygments
    toc: yes
    dev: pdf
vignette: >
  %\VignetteIndexEntry{Selection quantification} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\usepackage[utf8]{inputenc}
---

BASELINe quantifies selection pressure by calculating the posterior probability 
density function (PDF) based on observed mutations compared to expected 
mutation rates derived from an underlying SHM targeting model. Selection is 
quantified via the following steps:

1. Load a Change-O tab-delimited database file.
2. Calculate the selection scores and group by relevant fields for comparison.
3. Plot and compare selection scores of different groups of sequences.

Load Change-O data
--------------------------------------------------------------------------------
A small example Change-O tab-delimited database file is included in the `shm` 
package. The example dataset consists of a subset of Ig sequencing data from an 
influenza vaccination study (Laserson et al., PNAS, 2014). The data include 
sequences from multiple time-points before and after the subject received an 
influenza vaccination. Quantifying selection requires the following fields 
(columns) to be present in the Change-O file: `SEQUENCE_IMGT` and 
`GERMLINE_IMGT_D_MASK`.

```{r, eval=TRUE, warning=FALSE, message=FALSE}
library(alakazam)
library(shm)

# Load Change-O file
file <- system.file("extdata", "Influenza.tab", package="shm")
db <- readChangeoDb(file)
```

Calculate selection scores and group by relevant fields for comparison
--------------------------------------------------------------------------------
The first step to calculate selection is to call `getClonalConsensus` to 
collapse sequences by the `CLONE` column, if the data has been divided into 
clonal groups. 

```{r, eval=TRUE, warning=FALSE, results="hide"}
# Collapse clonal groups into single sequences
db_clone <- getClonalConsensus(db, regionDefinition=IMGT_V_NO_CDR3, 
                               collapseByClone=TRUE, nproc=3)
```

Following construction of an effective sequence for each clone, the observed
and expected mutation counts are calculated for each sequence of the 
`CLONAL_CONSENSUS_SEQUENCE` column adding in the previous step. 
`getObservedMutations` is used to calculate the number of observed mutations 
and `getExpectedMutationFreq` calculates the expected frequency of mutations. 
The underlying targeting model for calculating expectation can be specified 
using the `targetingModel` parameter. In the example below, the default 
`HS5FModel` is used. Column names for sequence and germline sequence may
also be passed in as parameters if they differ from Change-O defaults. 

Mutations are counted by these functions separately for complementarity 
determining (CDR) and framework (FWR) regions. If Ig sequences are not gapped 
according to IMGT numbering, the region boundaries may be defined using the 
`regionDefinition` parameter. There are four built-in region definitions 
in the `shm` package: 

*  `IMGT_V` is all regions in the V-segment grouped as either CDR or FWR.
*  `IMGT_V_BY_REGIONS` is all CDR and FWR regions in the V-segment treated as 
   individual regions.
*  `IMGT_V_NO_CDR3` (the default value of `regionDefinition`) is all regions in 
   the V-segment, excluding CDR3, grouped as either CDR or FWR.
*  `IMGT_V_BY_REGIONS_NO_CDR3` is all regions in the V-segment, excluding CDR3, 
   treated as individual regions. 

```{r, eval=TRUE, warning=FALSE, results="hide"}
# Count observed mutations and add OBSERVED columns to db_clone
db_obs <- getObservedMutations(db_clone, sequenceColumn="CLONAL_CONSENSUS_SEQUENCE",
                               regionDefinition=IMGT_V_NO_CDR3, nproc=3)
# Calculate expected mutations and add EXPECTED columns to db_obs
db_exp <- getExpectedMutationFreq(db_obs, sequenceColumn="CLONAL_CONSENSUS_SEQUENCE",
                                      targetingModel=HS5FModel,
                                      regionDefinition=IMGT_V_NO_CDR3, nproc=3)
````

The counts of observed and expected mutations can be combined to test for selection 
using `calcBaseline`. The statistical framework used to test for selection based 
on mutation counts can be specified using the `testStatistic` parameter. If the 
previous three functions have not yet been called on the input `data.frame` 
(as demonstrated by having columns beginning with `OBSERVED` and `EXPECTED`), 
this function will calculate observed and expected mutations after collapsing by 
the `CLONE` column.

```{r, eval=TRUE, warning=FALSE, results="hide"}
# Calculate selection scores using the output from getExpectedMutationFreq
baseline <- calcBaseline(db_exp, testStatistic="focused", nproc=3)

# Subset the original data to two time-points and switched isotypes
db_sub <- subset(db, CPRIMER %in% c("IGHA", "IGHG") & 
                     BARCODE %in% c("RL013", "RL018"))
# Calculate selection scores from scratch on subset
baseline_sub <- calcBaseline(db_sub, testStatistic="focused", 
                             regionDefinition=IMGT_V_NO_CDR3, nproc=3)
```

Selection scores can be used to compare selection pressures between different 
samples. In the example influenza dataset, the `BARCODE` field corresponds to 
samples taken at different timepoints before and after an influenza vaccination. 
Additionally, the `CPRIMER` field specifies the isotype of the sequence. The 
`groupBaseline` function convolves the Baseline PDFs of individual sequences 
(or clones) to get a combined PDF. The field(s) by which to group the sequences are 
specified with the `groupBy` parameter. The `groupBaseline` function automatically
calls `summarizeBaseline` to generate summary statistics based on the requested 
groupings, and populates the `stats` slot of the input `Baseline` object with the 
number of sequences with observed mutations for each region, selection scores, 95% 
confidence intervals, and p-values for each grouping. 

```{r, eval=TRUE, warning=FALSE, results="hide"}
# Combine selection scores by time-point
baseline_one <- groupBaseline(baseline, groupBy=c("BARCODE"))

# Combine selection scores by time-point and isotype
baseline_two <- groupBaseline(baseline, groupBy=c("BARCODE", "CPRIMER"))
baseline_sub <- groupBaseline(baseline_sub, groupBy=c("BARCODE", "CPRIMER"))
```

Plot and compare selection scores of different groups of sequences
--------------------------------------------------------------------------------
`plotBaselineSummary` plots the mean and confidence interval of selection scores 
for the given groups. The `idColumn` argument specifies the field that contains 
identifiers of the groups of sequences. If there is a secondary field by which 
the sequences are grouped, this can be specified using the `groupColumn`. This 
secondary grouping can have a user-defined color palette passed into 
`groupColors` or can be separated into facets by setting the `facetBy="group"`. 
The `subsetRegions` argument can be used to visualize selection of specific 
regions. Several examples utilizing these different parameters are provided 
below.

```{r, eval=TRUE, warning=FALSE}
# Plot mean and confidence interval by time-point
plotBaselineSummary(baseline_one, "BARCODE")

# Plot selection scores by time-point and isotype for only CDR
plotBaselineSummary(baseline_two, "BARCODE", "CPRIMER", subsetRegions="CDR")

# Plot only two time-points and recolor isotypes
group_colors <- c("IGHM"="darkorchid", "IGHD"="firebrick", 
                  "IGHG"="seagreen", "IGHA"="steelblue")
plotBaselineSummary(baseline_sub, "BARCODE", "CPRIMER", groupColors=group_colors)

# Group by CDR/FWR and facet by isotype
plotBaselineSummary(baseline_two, "BARCODE", "CPRIMER", facetBy="group")
```

`plotBaselineDensity` plots the full `Baseline` PDF of selection scores for the 
given groups. The parameters are the same as those for `plotBaselineSummary`.
However, rather than plotting the mean and confidence interval, the full density 
function is shown.

```{r, eval=TRUE, warning=FALSE}
# Plot selection PDFs for a subset of the data
group_colors <- c("RL013"="steelblue", "RL018"="firebrick")
plotBaselineDensity(baseline_sub, "CPRIMER", "BARCODE",
                    groupColors=group_colors, sigmaLimits=c(-3, 3))
```