---
title: 'SHM: Tuning clonal assignment thresholds by calculating distance to nearest
  neighbor'
author: "Namita Gupta"
date: '`r Sys.Date()`'
output:
  pdf_document:
    dev: pdf
    fig_height: 4
    fig_width: 7.5
    highlight: pygments
    toc: yes
  html_document:
    fig_height: 4
    fig_width: 7.5
    highlight: pygments
    theme: readable
    toc: yes
geometry: margin=1in
fontsize: 11pt
vignette: >
  %\VignetteIndexEntry{Distance to nearest neighbor} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\usepackage[utf8]{inputenc}
---

Estimating the optimal distance threshold for partitioning clonally related 
sequences is accomplished by calculating the distance from each sequence in the 
data set to its nearest neighbor and finding the break point in the resulting 
bi-modal distribution that separates clonally related from unrelated sequences. 
This is done via the following steps:

1. Load a Change-O tab-delimited database file.
2. Calculate the nearest neighbor distances for each sequence.
3. Generate a histogram of the nearest neighbor distances and inspect for the 
   threshold separating the two modes.

Load Change-O data
--------------------------------------------------------------------------------
A small example Change-O tab-delimited database file is included in the `shm` 
package. Calculating the nearest neighbor distances requires the following 
fields (columns) to be present in the Change-O file: `V_CALL`, `J_CALL`, 
`JUNCTION_LENGTH`, and `JUNCTION`.

```{r, eval=TRUE, warning=FALSE, message=FALSE}
library(alakazam)
library(shm)

# Load Change-O file
file <- system.file("extdata", "Influenza.tab", package="shm")
db <- readChangeoDb(file)
```

Calculate nearest neighbor distances
--------------------------------------------------------------------------------
The function for calculating distance between every sequence and its nearest
neighbor takes a few parameters to adjust how the distance is measured. If a 
genotype has been inferred using the methods in the `tigger` package, and a 
`V_CALL_GENOTYPED` field has been added to the database, then this column may be 
used instead of the default `V_CALL` column by specifying the `vCallColumn` 
argument. This will allows the more accurate V call from `tigger` to be used for 
grouping of the sequences. Furthermore, for more leniency toward ambiguous 
V(D)J segment calls, the parameter `first` can be set to `FALSE`. Setting 
`first=FALSE` will use the union of all possible genes to group sequences, rather 
than the first gene in the field. The `model` parameter determines which 
underlying SHM model is used to calculate the distance. The default model is 
`m1n`, an Ig-specific single nucleotide model similar to a 
transition/transversion model (Smith et al, J Immunol, 1996). Other options 
include nucleotide Hamming distance (`ham`), amino acid Hamming distance (`aa`), 
a 3-mer mouse model (`m3n`) (Smith et al, J Immunol, 1996), and a 5-mer model 
inferred from human data (`hs5f`) (Yaari et al, Front Immunol, 2013).

```{r, eval=TRUE, warning=FALSE}
# Use m1n model and normalize by junction length
dist_m1n <- distToNearest(db, model="m1n", first=FALSE, normalize="length")

# Use genotyped V assignments and 5-mer model
dist_hs5f <- distToNearest(db, vCallColumn="V_CALL_GENOTYPED", model="hs5f", 
                           first=FALSE)
```

Generate histogram
--------------------------------------------------------------------------------
The primary use of the distance to nearest calculation in the Change-O pipeline 
is to determine the optimal threshold for separating clonally related sequences 
(represented by sequences with "near"" neighbors) from singletons (sequences 
without "near"" neighbor), which show up as two modes in a histogram.

```{r, eval=TRUE, warning=FALSE, fig.width=7}
# Generate m1n histogram
library(ggplot2)
p1 <- ggplot() + theme_bw() + 
    ggtitle("Distance to nearest: m1n") + xlab("distance") +
    geom_histogram(data=dist_m1n, aes(x=DIST_NEAREST), binwidth=0.025, 
                   fill="steelblue", color="white") +
    geom_rug(aes(x=seq(0.0, 0.5, 0.1)), color="firebrick")
plot(p1)
```

In this example, the length normalized `m1n` model distance threshold would be 
set to a value near 0.1.

```{r, eval=TRUE, warning=FALSE, fig.width=7}
# Generate hs5f histogram
p2 <- ggplot() + theme_bw() + 
    ggtitle("Distance to nearest: hs5f") + xlab("distance") +
    geom_histogram(data=dist_hs5f, aes(x=DIST_NEAREST), binwidth=1, 
                   fill="steelblue", color="white") +
    geom_rug(aes(x=seq(0, 15, 1)), color="firebrick")
plot(p2)

# Zoom in to find threshold
p3 <- p2 + xlim(c(0, 15)) + scale_y_sqrt()
plot(p3)
```

In this example, the unnormalized `hs5f` model distance threshold would be 
set to a value near 5.